\textbf{Efficient Implementation Strategies for Block Ciphers on ARMv8}\\ {\footnotesize Bachelorarbeit}

### 2023-05-18 Paper/chapters/appendix.tex
rename files, add implementation sample code

\lstinputlisting[language=c, caption={\texttt{gift\_table.h}}]{../impl/table/gift_table.h}
\lstinputlisting[language=c, caption={\texttt{gift\_table.c}}]{../impl/table/gift_table.c}

@@

\lstinputlisting[language=c, caption={\texttt{gift\_vec\_sbox.h}}]{../impl/vector/gift_vec_sbox.h}
\lstinputlisting[language=c, caption={\texttt{gift\_vec\_sbox.c}}]{../impl/vector/gift_vec_sbox.c}

@@

\lstinputlisting[language=c, caption={\texttt{gift\_vec\_sliced.h}}]{../impl/vector/gift_vec_sliced.h}
\lstinputlisting[language=c, caption={\texttt{gift\_vec\_sliced.c}}]{../impl/vector/gift_vec_sliced.c}

>>>>

\lstinputlisting[language=c, caption={\texttt{gift\_table.h}}]{../impl/gift/table.h}
\lstinputlisting[language=c, caption={\texttt{gift\_table.c}}]{../impl/gift/table.c}

@@

\lstinputlisting[language=c, caption={\texttt{gift\_vec\_sbox.h}}]{../impl/gift/vec_sbox.h}
\lstinputlisting[language=c, caption={\texttt{gift\_vec\_sbox.c}}]{../impl/gift/vec_sbox.c}

@@

\lstinputlisting[language=c, caption={\texttt{gift\_vec\_sliced.h}}]{../impl/gift/vec_sliced.h}
\lstinputlisting[language=c, caption={\texttt{gift\_vec\_sliced.c}}]{../impl/gift/vec_sliced.c}

### 2023-05-18 Paper/chapters/implementation-strategies.tex
add macros to paper
rename files, add implementation sample code

% TODO change shr/shl implementation to only use vector registers
One caveat of this approach is the fact that NEON registers cannot be shifted
in their entirety due to the fact bits are not able to cross lanes. This leads
to the problem of being able to shift at most two lanes of 64 bits at once. We
thus need to implement the \texttt{shr(V,n)} and \texttt{shl(V,n)} operations
on our own. This can be done by extracting and shifting the overflow $ov$ out
of $V=b||a$, shifting the lanes individually and finally ORing the overflow
bits to the corresponding vector element.

\begin{alignat*}{2}
    &\text{shl}(V,n): \\
    &\qquad ov&&=V[0]\land(\lnot(\texttt{0xffffffffffffffff}\gg_n))\gg_{63-n} \\
    &\qquad V[0]&&=V[0]\gg_n \\
    &\qquad V[1]&&=(V[1]\gg_n)\lor ov
\end{alignat*}

The following operations group all $i$-th bits of the messages $a,c,\dots,o$
into bytes and puts these into the lower half of the registers $V_{i\bmod 8}$.
The same is done for messages $b,d,\dots,p$, only differing in that the bytes
are put into the upper half of the registers.

@@

We choose to encrypt 16 plaintext blocks at once so we can fill 16 vector
registers with register $V_i$ containing all $i$-th bytes. This lends itself well
to a NEON implementation since S-boxes as well as the FL layer, Feistel round
and packing/unpacking functions can be implemented efficiently.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{Figures/byteslicing.pdf}
    \caption{Camellia byte-sliced packing function with 32 \texttt{TBL} operations.
    Curly bracets represent a 4-element vector array being used as source registers.}
    \label{fig:byteslicing}
\end{figure}

@@

A byte-sliced implementation strategy for the S-box on ARMv8 can be derived
from already existing x86-optimized implementations utilizing the AES-NI
advanced encryption standard instruction set \cite{bcfastimplx86:2013}. NEON
itself possesses cryptographic extensions for finite field arithmetic and AES
as well as SHA calculations. These can be used to produce an accelerated
Camellia implementation due to the algebraic similarity of the AES- and
Camellia S-boxes.

@@

% TODO add graphic here

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{Figures/camelliaround.pdf}
    \caption{Camellia byte-sliced Feistel round function}
\end{figure}

@@

The two functions $FL,FL^{-1}$ can be implemented in a straightforward manner
once 1-bit left rotation has been defined for the bytesliced representation.
This can be facilitated by extracting the highest bit of each byte, shifting it
all the way to the right and storing it as overflow $o_i$. $o_i$ is then added
to the left-shifted value of byte $(i + 4) (\bmod 4)$ for the final result.

% TODO add graphic here

>>>>

% TODO change shr/shl implementation to only use vector registers
One caveat of this approach is the fact that NEON registers cannot be shifted
in their entirety due to the fact bits are not able to cross lanes. This leads
to the problem of being able to shift at most two lanes of 64 bits at once. We
thus need to implement the \texttt{shr(V,n)} and \texttt{shl(V,n)} operations
on our own. This can be done by extracting and shifting the overflow $ov$ out
of $V=V[1]||V[0]$, shifting the lanes individually and finally ORing the
overflow bits to the corresponding vector element.

\begin{alignat*}{2}
    &\text{shl}(V,n): \\
    &\qquad ov&&=V[0]\gg_{64-n} \\
    &\qquad V[0]&&=V[0]\ll \\
    &\qquad V[1]&&=(V[1]\ll)\lor ov
\end{alignat*}

The following operations group all $i$-th bits of the messages $a,c,\dots,o$
into bytes and put these into the lower half of the registers $V_{i\bmod 8}$.
The same is done for messages $b,d,\dots,p$, only differing in that the bytes
are put into the upper half of the registers.

@@

We choose a bytesliced representation by encrypting 16 plaintext blocks at once
so we can fill 16 vector registers with register $V_i$ containing all $i$-th
bytes. This lends itself well to a NEON implementation since S-boxes as well as
the FL layer, Feistel round and packing/unpacking functions can be implemented
efficiently.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{Figures/byteslicing.pdf}
    \caption{Camellia bytesliced packing function with 32 \texttt{TBL} operations.
    Curly brackets represent a 4-element vector array being used as source registers.}
    \label{fig:byteslicing}
\end{figure}

@@

A bytesliced implementation strategy for the S-box on ARMv8 can be derived
from already existing x86-optimized implementations utilizing the AES-NI
advanced encryption standard instruction set \cite{bcfastimplx86:2013}. NEON
itself possesses cryptographic extensions for finite field arithmetic and AES
as well as SHA calculations. These can be used to produce an accelerated
Camellia implementation due to the algebraic similarity of the AES- and
Camellia S-boxes.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{Figures/camelliaround.pdf}
    \caption{Camellia bytesliced Feistel round function}
\end{figure}

@@

The two functions $FL,FL^{-1}$ can be implemented in a straightforward manner
once 1-bit left rotation has been defined for the bytesliced representation.
This can be facilitated by extracting the highest bit of each byte, shifting it
all the way to the right and storing it as overflow $ov_i$. $ov_i$ is then added
to the left-shifted value of byte $(i + 4) (\bmod 4)$ for the final result.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\textwidth]{Figures/camellia_bytesliced_rol.pdf}
    \caption{Camellia bytesliced 1-bit left rotate}
\end{figure}

### 2023-05-18 Paper/chapters/implementation.tex
rename files, add implementation sample code

\lstinputlisting[language=c, firstline=77, lastline=87]{../impl/table/gift_table.c}
\lstinputlisting[language=c, firstline=89, lastline=101]{../impl/table/gift_table.c}

Lots of operations require extraction of a certain number of consecutive bits,
usually referred to as a bitfield. Indices for table lookups are generally
attained by right-shifting a larger value stored in a register, then applying
an AND operation to get the lowest $n$ bits and finally writing the result into
the beginning of another register. Due to its usefulness, this operation is
implemented as \texttt{UBFX} for an unsigned bitfield extract and can be used
to implement S-box extraction for subperm lookups which would otherwise take
two or three instructions.

@@

\lstinputlisting[language=c, firstline=92, lastline=95]{../impl/vector/gift_vec_sbox.c}
\lstinputlisting[language=c, firstline=102, lastline=117]{../impl/vector/gift_vec_sbox.c}
\lstinputlisting[language=c, firstline=197, lastline=212]{../impl/vector/gift_vec_sbox.c}

\subsection{Bitslicing}
\subsubsection{Shifts for swapmove}

@@

\begin{lstlisting}{language=c, caption={Shift and swapmove}}

/*
uint8x16_t shl(const uint8x16_t v, const int n) */
#define shl(_a, v, n)                                                       \
{                                                                           \
        uint64x2_t _overflow = vshrq_n_u64(v, 64 - n);                      \
        _overflow = vextq_u64(vdupq_n_u64(0x0), _overflow, 1);              \
        _a = vorrq_u8(vshlq_n_u64(v, n), _overflow);                        \
}

/*
uint8x16_t shr(const uint8x16_t v, const int n) */
#define shr(_a, v, n)                                                       \
{                                                                           \
        uint64x2_t _overflow = vshlq_n_u64(v, 64 - n);                      \
        _overflow = vextq_u64(_overflow, vdupq_n_u64(0x0), 1);              \
        _a = vorrq_u8(vshrq_n_u64(v, n), _overflow);                        \
}

/* implemented as a macro so we can use vshlq_n_u8 with variable n
void gift_64_vec_sliced_swapmove(uint8x16_t *restrict a, uint8x16_t *restrict b,
                                 const uint8x16_t m, const int n) */
#define gift_64_vec_sliced_swapmove(a, b, m, n)                             \
{                                                                           \
        uint8x16_t _a;                                                      \
        shr(_a, a, n);                                                      \
        const uint8x16_t _t = vandq_u8(veorq_u8(_a, b), m);                 \
        b = veorq_u8(b, _t);                                                \
        shl(_a, _t, n);                                                     \
        a = veorq_u8(a, _a);                                                \
}
\end{lstlisting}

@@

@@

Round keys are loaded from memory a few instructions before they are needed;
assuming all round keys are stored in the L1 cache, loading a
floating-point/vector register takes 5 cycles. After the load has been
dispatched to the load execution unit in line \ref{lst:floatload}, the
processor happily continues chugging away at the instruction stream by issuing
\texttt{tbl} and \texttt{mvn} $\mu$ops to other execution units.

@@

>>>>

\lstinputlisting[language=c, caption={GIFT-64 table subperm}, firstline=77, lastline=87]{../impl/gift/table.c}
\lstinputlisting[language=c, caption={GIFT-64 table encrypt}, firstline=89, lastline=101]{../impl/gift/table.c}

Lots of operations require extraction of a certain number of consecutive bits,
usually referred to as a bitfield. Indices for table lookups are generally
attained by right-shifting a larger value stored in a register, then applying
an AND operation to get the lowest $n$ bits and finally writing the result into
the beginning of another register. Due to its usefulness, this operation is
implemented as \texttt{UBFX} for an unsigned bitfield extract and can be used
to implement S-box extraction for subperm lookups which would otherwise take
two or three instructions. Interestingly, the AArch64 instruction set makes
heavy use of instruction aliasing. The logical shift left instruction
\texttt{lsl} for example is an alias of \texttt{UBFX} which itself is an alias
for \texttt{UBFM}.

Another keyword aiding in optimization is \texttt{restrict} which can be used
for pointer and array function parameters; the programmer can add this keyword
to parameters to tell the compiler they are never aliased by any other pointers
which allows the compiler to rearrange instructions and eliminate loads.

@@

\lstinputlisting[language=c, caption={\texttt{vperm} S-box}, firstline=95, lastline=98]{../impl/gift/vec_sbox.c}
\lstinputlisting[language=c, caption={\texttt{vperm} permutation}, firstline=105, lastline=120]{../impl/gift/vec_sbox.c}
\lstinputlisting[language=c, caption={\texttt{vperm} encrypt function}, firstline=200, lastline=215]{../impl/gift/vec_sbox.c}

\subsection{Bitslicing}

\subsubsection{A note on data storage}

NEON only provides 32 vector registers. While this is more than the 16 YMM
registers offered by Intel's AVX-256 vector extension, it is not enough to
accompany all 28 round keys for GIFT-64 plus the 16 registers representing
cipher state at once. Because loading and storing single registers is
inefficient, data is represented using the vector array type
\texttt{uint8x16x4\_t}. Loads and stores are then assembled such that the whole
array is loaded or stored in a single instruction. Loading a single vector from
memory for example has a latency of 5 cycles while loading four vectors into a
vector array can be done in 8 cycles which results in an amortized cost of 2
cycles per vector. Vectors are grouped into vector arrays as often as possible
to reduce the number of necessary loads and stores.

\subsubsection{Shifts for swapmove}

@@

\lstinputlisting[language=c, caption={Bitsliced GIFT swapmove and shift macros}, firstline=50, lastline=79]{../impl/gift/vec_sliced.c}

@@

\lstinputlisting[language=c, caption={Bitsliced GIFT S-box}, firstline=157, lastline=180]{../impl/gift/vec_sliced.c}
\lstinputlisting[language=c, caption={Bitsliced GIFT permutation}, firstline=205, lastline=216]{../impl/gift/vec_sliced.c}

@@

Round keys are loaded from memory a few instructions before they are needed;
assuming all round keys are stored in the L1 cache, loading a
floating-point/vector register takes 5 cycles. After the load has been
dispatched to the load execution unit in line \ref{lst:floatload}, the
processor continues processing the instruction stream by issuing \texttt{tbl}
and \texttt{mvn} $\mu$ops to other execution units.

@@

\section{Camellia}
\subsection{Optimized non-SIMD implementation}

An optimized non-SIMD implementation based on the platform-independent
techniques presented in the original paper is relatively easy to achieve since
all functional components are already well defined. One thing to take note of
however is the fact that the specification is based on a big endian
representation while ARMv8 is a little endian machine. The problem of
endianness manifests itself whenever a memory-register interaction takes place,
i.e. for loads and stores. Arithmetic on a register is unaffected since a
register is always treated as one large number with no conception of addresses.

We will store input data and arrays in Camellia byte-order such that a lower
array element actually represents a higher numerical value, i.e. the numerical
value of

\begin{center}
    \texttt{\{ 0x0123456789abcdefUL, 0xfedcba9876543210UL \}}
\end{center}

is equal to

\begin{center}
    \texttt{0x0123456789abcdeffedcba9876543210}
\end{center}

which is equal to the byte string

\begin{center}
    \texttt{01 23 45 67 89 ab cd ef fe dc ba 98 76 54 32 10}
\end{center}

This allows us to change the endianness of the implementation while keeping
input in the original Camellia byte order. A change in implementation
endianness manifest itself for example in table lookups where, according to
Camellia specification, the first byte is expected to be the most significant
byte:

\lstinputlisting[language=c, caption={Camellia optimized SP function with reverse byte order}, firstline=29, lastline=45]{../impl/camellia/spec_opt.c}

\subsection{Bytesliced implementation}

\subsubsection{Key generation}

A possible approach to storing keys is storing them in bytesliced format so
they can later be loaded from memory and used directly since we only have 32
vector registers in total and half of them are already occupied by the current
cipher state. This can be implemented by first generating 64 bit round keys and
then filling register $V_i$ with the $i$-th byte by means of
\texttt{vdupq\_n\_u8} for each key.

\lstinputlisting[language=c, caption={Bytesliced Camellia round key generation}, firstline=237, lastline=266]{../impl/camellia/bytesliced.c}

Another way to implement key generation is not to store the keys bytesliced,
but rather to store the 64 bit values and create necessary vector registers on
the fly. This saves 3120 bytes of memory and could show some usefulness for
memory-constrained environments, but benchmarking shows a performance drop of
about $3.6\%$. This is due to the fact that \texttt{vdupq\_n\_u8} has a latency
of 8 cycles and a throughput of only 1, making it a great deal slower than
simple vector array loads.

\subsubsection{F-function}

The s-function implements matrix multiplication with a given pre- and
postfilter using the aforementioned strategy.

\lstinputlisting[language=c, caption={Bytesliced Camellia S-box}, firstline=40, lastline=65]{../impl/camellia/bytesliced.c}

Just as before we need to invert the byte order since the first vector $V_0$
contains all least significant bytes, but the Camellia specification places the
most significant bytes at the beginning.

\lstinputlisting[language=c, caption={Bytesliced Camellia F-function}, firstline=67, lastline=109]{../impl/camellia/bytesliced.c}
